---
title: "Assignment 4: Exploratory Data Analysis"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Overview

Due: **Sunday, 9/28 11:59pm**

In this assignment, you will practice Exploratory Data Analysis (EDA) by making plots from a **Kaggle dataset of your choice**.

You will submit **at least 4 different plot types**, with short written interpretations for each.

Part 1 is worth a total of 1 point, and Part 2 is worth 4 points (1 point for each plot & interpretation).

## Part 1: Get Your Data

1.  Go to [Kaggle Datasets](https://www.kaggle.com/datasets), create an account, and choose a dataset that interests you. Ensure your data has:
    1.  At least 4 variables
    2.  At least one categorical and one continuous variable
    3.  I recommend data that contains at least 50-100 observations
    4.  Tip: Open the CSV on Kaggle first and skim the columns to check that it meets these requirements
2.  Download the dataset as a CSV file.
3.  Place the CSV in your class R project data folder.
4.  Read it into R with `read_csv()`.

```{r, message=FALSE}
library(tidyverse)

mydata <- read_csv("data/VNL2023.csv")

```

### Describe your data

Data author: Yeganeh Bavafa

Link: <https://www.kaggle.com/datasets/yeganehbavafa/vnl-men-2023/data>

Description: He collected and sorted a dataset about volleyball players in Volleyball National League in 2023. In each factor, he have used the average of the player during a game, which was the author of errors, points and percentage of positive performance of that player in each section.

-   Position Feature represent the real position of each player

1.  OH : outside hitter

2.  OP : Opposite hitter

3.  MB : Middle blocker

4.  S : Setter

5.  L : Libero

-   Attack : A player's overall average during each game in the offense factor

-   Block : A player's overall average during each game in the defense factor.\
    Defense on the net includes direct points, errors and touching the ball without changing points.

-   Serve : Except for the libero player, the rest of the players serve during each turn to start the game. Each player's service average during the match is listed here.

-   Set : The setter is responsible for setting the players. But in special cases, the rest of the players cooperate in this matter, and in the data analysis, we will find out which position the players have the most participation in setting after the setter.

-   Set feature represent the average of successful sets, errors and attempts for each ball during a rally.

-   Dig : the average of digs, errors and receptions

-   Receive : Receive feature also represents the average of successful receptions, errors and attempts per mach.

### Preview your data

Use some of the strategies you've learned in this class or picked up elsewhere to summarize and quickly explore your data.

```{r}
library(dplyr)
View(mydata)
summary(mydata)

```

### Summary stats

Before you make plots, you must also:

-   Show basic summaries of your dataset, including:

    -   `count()` for categorical variables (to see category frequencies)

    -   `mean()`, `median()`, `sd()`, `min()`, `max()` for at least one numeric variable

```{r}
# See how many people in each position
mydata %>%
    count(Position)
# See how many peopke in each country
mydata %>%
  count(Country)
# See how many peopke in each country
mydata %>%
  count(Age)


# See the mean value of Age
mean(mydata$Age)

# See the mean value of Age each Position
mydata %>%
  group_by(Position) %>%
  summarise(mean_age = mean(Age), na.rm = TRUE)

#See the mean value of Attack in each country without Libero and Setter
mydata %>%
  filter(Position %in% c('OH', 'MB', "OP")) %>%
  group_by(Country) %>%
  summarise(mean_attack_country = mean(Attack, na.rm=TRUE)) %>%
  arrange(mean_attack_country)

#See the mean value of Attack in each age without Libero and Setter
mydata %>%
  filter(Position %in% c('OH', 'MB', "OP")) %>%
  group_by(Age) %>%
  summarise(mean_attack_age = mean(Attack, na.rm=TRUE)) %>%
  arrange(mean_attack_age)
  
#See the mean value of Block in each country without Libero 
mydata %>%
  filter(Position != 'L') %>%
  group_by(Country) %>%
  summarise(mean_block_country = mean(Block, na.rm=TRUE)) %>%
  arrange(mean_block_country)

#See the mean value of Block in each age without Libero and Setter
mydata %>%
  filter(Position %in% c('OH', 'MB', "OP")) %>%
  group_by(Age) %>%
  summarise(mean_block_age = mean(Block, na.rm=TRUE)) %>%
  arrange(mean_block_age)

# See the mean value of Set in each country
mydata %>%
  filter(Position == 'S') %>%
  group_by(Country) %>%
  summarise(mean_set_country = mean(Set, na.rm=TRUE)) %>%
  arrange(mean_set_country)

# See the mean value of Set in each age
mydata %>%
  filter(Position == 'S') %>%
  group_by(Age) %>%
  summarise(mean_set_age = mean(Set, na.rm=TRUE)) %>%
  arrange(mean_set_age)






# See the median of Attack of each Position
mydata %>%
  group_by(Position) %>%
  summarise(median_attack = median(Attack, na.rm = TRUE))

# See the median of Dig of each Position
mydata %>%
  group_by(Position) %>%
  summarise(median_dig = median(Dig, na.rm = TRUE)) %>%
  arrange(desc(median_dig))

# See the median value of Attack in each country without Libero and Setter
mydata %>%
  filter(Position %in% c('OH', 'MB', "OP")) %>%
  group_by(Country) %>%
  summarise(median_attack_country = median(Attack, na.rm=TRUE)) %>%
  arrange(median_attack_country)



# See the Satndard Deviation of Age
sd(mydata$Age)


# See the minimum of Age
min(mydata$Age)

# See the minimum rate of the Attack in each country without Libero and Setter
mydata %>%
  filter(Position %in% c('OH', 'MB', "OP")) %>%
  group_by(Country) %>%
  summarise(min_attack_country = min(Attack, na.rm = TRUE)) %>%
  arrange(min_attack_country)
  


# See the max value of Age
max(mydata$Age)
```

Write 2–3 sentences here: What do the category counts tell you? What does the numeric summary show about spread, center, or outliers?

The category counts show how players distributed. Outside Hitters are the most common position. And the numeric summaries reveal that setters (S) and liberos (L) tend to be older on average compared to attacking positions. When I see the Attack and Block by age, it shows wide variability, so it looks age doesn't reflect the their performance.

## Part 2:

Create **at least 4 plots**, each of a different type. At least one of your four plots must use different variables than the others.

Each plot should have:

-   A title and labeled axesy

-   y

Use the EDA Part 3 notes as your reference. Since this is an EDA assignment, your plots can be plain, but consider adding colors, themes, or interactivity for interpretability and visual appeal if you know how.

### Plot 1

```{r}
# insert code here

# try to see high attack, high block?

```

Interpretation:

### Plot 2

```{r}
# insert code here

# see the set and attack
```

Interpretation:

### Plot 3

```{r}
# insert code here

# attack by position
```

Interpretation:

### Plot 4

```{r}
# insert code here

# Receive and Attack
```

Interpretation:

## Push to GitHub to Submit!

Assignment is due Sunday 9/28 at 11:59 pm.
