---
title: "Assignment 5: Substance Abuse Treatment Mapping"
subtitle: "STUDENT NAME"
author: "Amber Camp"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Overview

Due: **Tuesday, 10/7 11:59pm**

In this assignment, you will apply analytic and mapping skills on the TEDS-D dataset from SAMHSA.

Refer to the following files for reference, though some of what is asked of you in this assignment will require logic beyond what was discussed here:

-   week06-treatment-mapping-part1.qmd

-   week06-treatment-mapping-part2.qmd

Total time estimate: Two hours

## Load Packages and Data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(arrow) # lets us read/write parquet and feather files
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(naniar) # tools for exploring and visualizing missing data
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation

```

```{r}
teds_d <- read_parquet(here("data/teds_d_clean.parquet"))
```

## Question 1

Make an interactive map with `ggiraph` showing the percentage of completed treatments that end with no use at discharge. I am looking for a single interactive map as the final product.

Hint: Break this question down into bite-sized steps, like we did in Part 2. The `freq1_d` column contains information about use at discharge. Check your denominator carefully — should it be *all cases* or *completed cases*?

```{r}

```

## Question 2

Treatment outcomes can vary depending on the type of service patients receive and the length of stay in treatment.

-   Investigate how the percentage of treatments completed and the percentage of treatments that end with no use at discharge differ when you look at service type and length of stay.

-   Create at least three different visualizations to help answer this question.

-   At least one visualization should focus on service type and at least one should focus on length of stay. A third visualization should compare the two.

-   After making your visualizations, write a short summary (3-ish sentences) that explains the main patterns you observe. Please use full, grammatical sentences.

```{r}

```

Summary: (write here)

## Push to GitHub to Submit!

Assignment is due Tuesday 10/7 at 11:59 pm.
