---
title: "Assignment 5: Substance Abuse Treatment Mapping"
subtitle: "Rei Miyashiro"
author: "Rei Miyashiro"
date: "`r Sys.Date()`"
format: html
editor: visual
---

## Overview

Due: **Tuesday, 10/7 11:59pm**

In this assignment, you will apply analytic and mapping skills on the TEDS-D dataset from SAMHSA.

Refer to the following files for reference, though some of what is asked of you in this assignment will require logic beyond what was discussed here:

-   week06-treatment-mapping-part1.qmd

-   week06-treatment-mapping-part2.qmd

Total time estimate: Two hours

## Load Packages and Data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(arrow) # lets us read/write parquet and feather files
library(sf) # "simple features" for spatial data
library(urbnmapr) # access to US state/country shapefiles for mapping
library(naniar) # tools for exploring and visualizing missing data
library(ggiraph) # makes ggplot charts interactive
options(scipen = 99) # turns off scientific notation

```

```{r}
teds_d <- read_parquet(here("data/teds_d_clean.parquet"))
```

## Question 1

Make an interactive map with `ggiraph` showing the percentage of completed treatments that end with no use at discharge. I am looking for a single interactive map as the final product.

Hint: Break this question down into bite-sized steps, like we did in Part 2. The `freq1_d` column contains information about use at discharge. Check your denominator carefully — should it be *all cases* or *completed cases*?

```{r}
# Filter just reason = Treatment completed
completed <- teds_d %>%
  filter(reason == "Treatment completed")

# Calculate no use cases / total Treatment completed cases in each state
percent_completed_no_use_by_state <- teds_d %>%
  group_by(stfips) %>% 
  summarize(total_cases = n(), # this grabs the total number of obvs
    no_use_cases = sum(freq1_d == "no use", na.rm = TRUE)) %>% # filter "no use" cases from freq1_d
  mutate(percentage_no_use_completed = (no_use_cases / total_cases) * 100) # basic math :)

percent_completed_no_use_by_state %>% 
 select(percentage_no_use_completed)
summary(percent_completed_no_use_by_state)

# Need to bring the state map 
states_map <- get_urbn_map(map = "states", sf = TRUE)
head(states_map, 10)

# change the format to combine state_map and percent_completed_no_use_by_state
percent_completed_no_use_by_state <- percent_completed_no_use_by_state %>%
  mutate(state_fips = sprintf("%02d", stfips)) 
view(percent_completed_no_use_by_state)
view(states_map)

# join the data 
percent_completed_no_use_by_state_map <- full_join(percent_completed_no_use_by_state,
                          states_map,
                          by = "state_fips") # remember the index we prepared above?

# Plot the map
# catch-all bin for NA values and assign it a category name
extra_bin <- percent_completed_no_use_by_state_map %>% 
  mutate(percentage_bin = cut(percentage_no_use_completed, breaks=c(0, 10, 20, 30, 40, 50, 60, 70)),
         percentage_bin = fct_explicit_na(percentage_bin, na_level = "No Data"))


# Build interactrive map
no_use_map <- ggplot(extra_bin) +
  geom_sf_interactive(
    aes(geometry = geometry,
      fill = percentage_bin,
      tooltip = paste("State:", state_abbv,
                      "<br>FIPS:", stfips,
                      "<br>No use at Discharge:", round(percentage_no_use_completed,2), "%")), 
    color = "#ffffff", size = 0.25) +
  labs(fill = "% of Completed Cases with No Use at Discharge",
       title = "Treatment Episodes Ending with No Use at Discharge",
       subtitle = "TEDS-D Dataset (SAMHSA)") +
  scale_fill_viridis_d(option = "mako", direction = -1) + 
  coord_sf(datum = NA) +
  theme_minimal() +
  theme(panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.text = element_text(size = 4),
    legend.title = element_text(size = 5),
    strip.text = element_text(size = 4))

girafe(ggobj = no_use_map)
```

## Question 2

Treatment outcomes can vary depending on the type of service patients receive and the length of stay in treatment.

-   Investigate how the percentage of treatments completed and the percentage of treatments that end with no use at discharge differ when you look at service type and length of stay.

-   Create at least three different visualizations to help answer this question.

-   At least one visualization should focus on service type and at least one should focus on length of stay. A third visualization should compare the two.

-   After making your visualizations, write a short summary (3-ish sentences) that explains the main patterns you observe. Please use full, grammatical sentences.

```{r}
# Services outcome
serv_outcomes <- teds_d %>%
  group_by(services_d) %>%    
  summarise(
    number_all = n(),
    total_completed = sum(reason == "Treatment completed"),
    percentage_completed_all = ((total_completed / number_all) * 100),
    total_no_use = sum(freq1_d == "no use"),
    percentage_no_use = ((total_no_use / number_all) * 100),
    percentage_no_use_completed = ((total_no_use / total_completed))
  )
# try to make a bar
ggplot(serv_outcomes, aes (x = services_d, y = percentage_no_use_completed)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "No Use at Discharge (of Completed) by Service Type",
       x = "Service Type",
       y = "Percent No Use (of Completed)") +
  geom_col(fill = "steelblue")





# lenght of stay outcome
los_outcomes <- teds_d %>%
  group_by(los) %>%    
  summarise(
    number_all = n(),
    total_completed = sum(reason == "Treatment completed"),
    percentage_completed_all = ((total_completed / number_all) * 100),
    total_no_use = sum(freq1_d == "no use"),
    percentage_no_use = ((total_no_use / number_all) * 100),
    percentage_no_use_completed = ((total_no_use / total_completed))
  )

# line 
ggplot(los_outcomes, aes(x = los, y = percentage_no_use_completed)) +
  labs(title = "No Use at Discharge (of Completed) by Length of Stay",
       x = "Length of Stay", y = "",
       fill = "No Use % (of Completed)") +
  geom_line()





#combine them
both_outcomes <- teds_d %>%
  group_by(services_d, los) %>%    
  summarise(
    number_all = n(),
    total_completed = sum(reason == "Treatment completed"),
    percentage_completed_all = ((total_completed / number_all) * 100),
    total_no_use = sum(reason == "Treatment completed" & freq1_d == "no use"),
    percentage_no_use_completed = ((total_no_use / total_completed))
  )

# line chart to see relationship them
ggplot(both_outcomes,
       aes(x = los, y = percentage_no_use_completed, color = services_d)) +
  geom_line() +
  labs(title = "No Use at Discharge (of Completed) by LOS, colored by Service",
       x = "Length of stay (days)", y = "% No use (of completed)", color = "Service") +
  theme_minimal()

# Using the facet to see each services
ggplot(both_outcomes,
       aes(x = los, y = percentage_no_use_completed, color = services_d)) +
  geom_line() +
  labs(title = "No Use at Discharge (of Completed) by LOS, colored by Service",
       x = "Length of stay (days)", y = "% No use (of completed)", color = "Service") +
  theme_minimal() +
facet_wrap(~ services_d)

```

Summary: Completion rates and no-use-at-discharge outcomes vary by service type. For most services types, the line slopes upward as Length of stay increase, so the patients stays longer in treatment, the more likely they are to finish treatment without using substances. Some services show steeper improvements while some are flatter (Ambulatory, detoxification and unknown) but the overall pattern is a slightly positive relationship between length of stay and completed treatment without substances.

## Push to GitHub to Submit!

Assignment is due Tuesday 10/7 at 11:59 pm.
